The Dynamics of Anthropomorphic Effects in Human-Robot Interaction

### 2014-03-19 paper.tex
refined intro

The essence of anthropomorphism is the perception of human-like characteristics
in either real or imagined non-human agents. In Human-Robot Interaction (HRI),
anthropomorphism is a commonly observed phenomenon: people tend to ascribe
human-like characteristics to robots, and they are willing to treat and react to
robots as if they were human-social counterparts. The HRI and social robotics
literature is full of various kinds of studies showing how humans readily
interact in a social fashion with robots -- despite the fact that robots are not
humans. In our own studies on HRI analysis in real-world settings, we also
observed anthropomorphism. However, we found qualitative differences and
dynamics over time in how participants anthropomorphized and related to the
robots used in our studies. More concretely, we found that anthropomorphism was
not only based on how much the robot resembles a human (anthropomorphic design)
but that it was determined by multiple factors, including individual
characteristics of the human user, and the context of use, for instance. We also
found that anthropomorphism changed over time with growing experience of the
user with the system. By trying to understand these variances and dynamics in
anthropomorphism, we reviewed related work from the HRI domain and social
robotics. Despite the fact that there is a vast amount of literature on
anthropomorphism in HRI, we were surprised to find that other studies do not
describe similar variances or dynamics. Instead, we came across diverse and
partially conflicting understandings of anthropomorphism. We also had the
impression that the reported data were sometimes over-interpreted (probably due
to the confusion of what exactly anthropomorphism is) or obtained in settings
and assessed in ways that we believe do not give enough consideration to the
complex nature of anthropomorphism. However, when reviewing literature from
social psychology, cognitive science, and neuroscience, we found support for the
variances and dynamics in anthropomorphism that we observed in our studies. As
HRI integrates many disciplines, we think that it is valid to use theories from
these related domains for interpreting our results and that it is relevant to
report on this integration.

In this article, we would like to share some critical but constructive
considerations on anthropomorphism in robotics and HRI. Thereby, we contribute
to a better understanding of this complex phenomenon related to robots. In
Section \ref{sec:anthropomorphism} we would first like to give an overview of
how anthropomorphism can be defined, and then, in Section \ref{sec:explanations}
present explanations to the question why people anthropomorphize, with a focus
on anthropomorphism as a social phenomenon in HRI and the trend to design
human-like robots. Then, based on these reviews of theories and concepts from
related domains, in Section \ref{sec:multi-factors} we propose that
anthropomorphism is determined and affected by multiple factors. We outline how
anthropomorphism as a multi-factor phenomenon can impact the user experience and
acceptance of robots in daily life environments. Next, in Section
\ref{sec:dynamics-model} we suggest that anthropomorphism changes over time, and
describe these changes in an initial model that we call ``Dynamics of
Anthropomorphism''. We illustrate our idea with empirical findings from several
HRI studies that we conducted in natural environments. In Section
\ref{sec:cognition-neuroscience} we present a cognitive and neuroscience
perspective on anthropomorphism to consolidate our considerations and the
observed dynamics. Next, Section \ref{sec:discussion} considers limitations in
our approach and we discuss existing and alternative ways to measure
anthropomorphism in HRI. Finally, in Section \ref{sec:conclusion} we conclude
this article.

>>>>

%{Intro when start with anthropomorphic design}
The Uncanny Valley hypothesis \citep{mori_uncanny_1970} proposes that human-like designs of objects and non-human agents evoke a positive feeling of affinity in the human observer / user, which in turn reflects the experience of emotional engagement with and feelings of empathy for the non-human agent \citep{cheetham_human_2011}. The positive effect of so-called \textbf{anthropomorphic form} is exploited in the design of artificial agents (\eg digital characters, robots) for several reasons. Amongst others, anthropomorphic design should serve to elicit this positive valence as it is predicted in the Uncanny Valley hypothesis, and it should further foster people's acceptance of these artificial agents, and enable effective (human-like and especially human-social) interaction. 
However, at a certain point as the non-human agent approaches but fails to closely resemble a human, the evoked valence in the human user suddenly turns into a negative uncanny feeling of what \cite{mori_uncanny_1970} describes as disgust and revulsion. This, in turn, might lead to rejection of the non-human but human-like agent.
The Uncanny Valley hypothesis predicts that with further increasing the human-likeness of the non-human agent such that it appears so realistic that it is not distinguishable from a human but perceived to be human, the effect is again a positive one, reflecting feelings and engagement that are similar as those evoked by a human-human interaction.
Of course, roboticists and developers try to avoid ``falling'' into this Uncanny Valley with their human-like designs of artificial agents. Consequently, there is a considerable research effort into better understanding and defining the boundaries of anthropomorphic design. However, findings have been inconsistent \citep{cheetham_human_2011}.

A phenomenon that is often studied related to the human-like design of robots is \textbf{anthropomorphism}. Besides a feeling of affinity, and empathy, anthropomorphism can be understood as being one of the resulting (and  more or less desired) effects of the anthropomorphic design of artificial agents (up to the first positive peak). The essence of anthropomorphism is the human user's perception of human-likeness in the non-human agent. Logically, anthropomorphism is said to be enhanced with increased human-like design of the agent. There are various other reasons and explanations for why people tend to anthropomorphize non-human agents. In general, as anthropomorphism emerges in a (real or imagined) interaction \citep{persson_anthropomorphism_2000} between at least two players (the human who anthropomorphizes and the non-human agent which is anthropomorphized) there are also two main perspectives on the phenomenon: an artifact-centered (design) and a human-centered (cognitive, psychological) one. The Uncanny Valley hypothesis can be categorized as being part of first, artifact-centered approach, to understand anthropomorphism.

We observed anthropomorphism in several short- and long-term human-robot interaction studies that we carried out with embodied robots, mostly in real-world settings.\footnote{Note that it was not the main purpose of these studies to analyze in detail the effect of human-likeness in a robot on the user's perception of and interaction with the robot.} There is not much research that investigates anthropomorphism based on real interaction between a human and a robot, and above all in a natural setting and over an extended period of time. 
Our data suggest that there are qualitative differences and dynamics over time in how participants relate to different types of robots. More concretely, we found that how human-like people perceived and interacted with the robot was not only based on how much the robot resembled a human (mostly in its behavioral cues)
but that this was determined by multiple factors, including individual
characteristics of the human user but also the context of use, for instance. Interestingly, our findings also suggest that anthropomorphism is likely to change over time with growing experience of the
user with the system.

Based on these observations and an extensive literature review, we would like to propose a model that accounts for variances and dynamics in anthropomorphism. This model of the ``Dynamics of Anthropomorphism'' considers variances in respect to multiple factors, including the human user, the anthropomorphic design of the non-human agent, the context of use, and the interaction experience (time).

Thereby, we would like to extend the current understanding of anthropomorphism in robotics and human-robot interaction.

%{Intro when start with anthropomorphism}
%The essence of anthropomorphism is the perception of human-like characteristics
%in either real or imagined non-human agents. In Human-Robot Interaction (HRI),
%anthropomorphism is a commonly observed phenomenon: people tend to ascribe
%human-like characteristics to robots, and they are willing to treat and react to
%robots as if they were human-social counterparts. The HRI and social robotics
%literature is full of various kinds of studies showing how humans readily
%interact in a social fashion with robots -- despite the fact that robots are not
%humans. In our own studies on HRI analysis in real-world settings, we also
%observed anthropomorphism. However, we found qualitative differences and
%dynamics over time in how participants anthropomorphized and related to the
%robots used in our studies. More concretely, we found that anthropomorphism was
%not only based on how much the robot resembles a human (anthropomorphic design)
%but that it was determined by multiple factors, including individual
%characteristics of the human user but also the context of use, for instance. We also
%found that anthropomorphism changed over time with growing experience of the
%user with the system. By trying to understand these variances and dynamics in
%anthropomorphism, we reviewed related work from the HRI domain and social
%robotics. Despite the fact that there is a vast amount of literature on
%anthropomorphism in HRI, we were surprised to find that other studies do not
%describe similar variances or dynamics. Instead, we came across diverse and
%partially conflicting understandings of anthropomorphism. We also had the
%impression that the reported data were sometimes over-interpreted (probably due
%to the confusion of what exactly anthropomorphism is) or obtained in settings
%and assessed in ways that we believe do not give enough consideration to the
%complex nature of anthropomorphism. However, when reviewing literature from
%social psychology, cognitive science, and neuroscience, we found support for the
%variances and dynamics in anthropomorphism that we observed in our studies. As
%HRI integrates many disciplines, we think that it is valid to use theories from
%these related domains for interpreting our results and that it is relevant to
%report on this integration.
%
%In this article, we would like to share some critical but constructive
%considerations on anthropomorphism in robotics and HRI. Thereby, we contribute
%to a better understanding of this complex phenomenon related to robots. In
%Section \ref{sec:anthropomorphism} we would first like to give an overview of
%how anthropomorphism can be defined, and then, in Section \ref{sec:explanations}
%present explanations to the question why people anthropomorphize, with a focus
%on anthropomorphism as a social phenomenon in HRI and the trend to design
%human-like robots. Then, based on these reviews of theories and concepts from
%related domains, in Section \ref{sec:multi-factors} we propose that
%anthropomorphism is determined and affected by multiple factors. We outline how
%anthropomorphism as a multi-factor phenomenon can impact the user experience and
%acceptance of robots in daily life environments. Next, in Section
%\ref{sec:dynamics-model} we suggest that anthropomorphism changes over time, and
%describe these changes in an initial model that we call ``Dynamics of
%Anthropomorphism''. We illustrate our idea with empirical findings from several
%HRI studies that we conducted in natural environments. In Section
%\ref{sec:cognition-neuroscience} we present a cognitive and neuroscience
%perspective on anthropomorphism to consolidate our considerations and the
%observed dynamics. Next, Section \ref{sec:discussion} considers limitations in
%our approach and we discuss existing and alternative ways to measure
%anthropomorphism in HRI. Finally, in Section \ref{sec:conclusion} we conclude
%this article.
